---
title: "Predicting body temperature distributions from TPC shapes"
description: "An exploration of how to predict body temperature distributions from thermal performance curve shapes and microclimate data."
date: 2024-03-05
categories: [model]
format:
  html:
    code-fold: true
    fig-align: "center"
    fig-dpi: 300
    lightbox: true
---

```{r setup}
#| code-summary: "Setup"
#| message: false
#| warning: false

# packages
library(tidyverse)
library(sn)
library(kableExtra)

```

## Context

I have been exploring if it is possible to predict the distribution of body temperatures that an ectotherm experiences in a given environment based on the shape of its thermal performance curve (TPC) and microclimate data. Below is a quick exploration of how to do so in R.

## Defining functions

First, I define a TPC function based on [Deutsch et al. (2008)](https://www.pnas.org/doi/10.1073/pnas.0709472105). This function uses T<sub>opt</sub>, CT<sub>min</sub>, and CT<sub>max</sub> to calculate the performance of an organism at a given body temperature. These parameters are often estimated from empirical data so it makes the whole thing very easily interpretable. 

```{r tpc-function}
#| code-summary: "Defining TPC function"
#| out-width: "75%"
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 3

# define TPC function
tpc <- function(temps, topt, ctmin, ctmax) {

  # initialize performance vector and define psig
  perf <- rep(NA, length(temps))
  psig <- -((ctmin - topt) / 4)

  # calculate performance
  for (i in seq_along(temps)){
    if (temps[i] < topt) {
      perf[i] <- exp(-((temps[i] - topt) / (2 * psig))^2)
    } else {
      perf[i] <- 1 - ((temps[i] - topt) / (topt - ctmax))^2
    }
  }

  # correct for negative values and return
  ifelse(perf < 0, 0, perf)

}

# example usage
temps <- seq(0, 40, by = 0.1)
topt <- 25
ctmin <- 10
ctmax <- 35
tpc_values <- tpc(temps, topt, ctmin, ctmax)
tibble(temps, tpc_values) |>
  ggplot(aes(x = temps, y = tpc_values)) +
  geom_line(linewidth = 2) +
  labs(x = "Body Temperature (°C)", y = "Performance") +
  theme_minimal()
```

Second, I define a function that generates a skew-normal distribution of body temperatures based on a mean, standard deviation, and skewness parameter. To do this I use two functions:

- `cp2dp()` from the `sn` package to convert the mean, sd, and skewness parameters to the location, scale, and shape parameters of the skew-normal distribution.
- `dsn()` from the `sn` package to calculate the density of the skew-normal distribution at a given set of body temperatures.

```{r create-tb-distribs}
#| code-summary: "Defining function to generate body temperature distributions"
#| out-width: "75%"
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 3

gen_tb_distrib <- function(temps, mean, sd, skew) {

  # stop if SD is negative or if skew is outside of -1 to 1
  if (sd < 0) stop("SD must be non-negative")
  if (skew <= -1 || skew >= 1) stop("Skew must be between -0.99 and 0.99")

  # estimate the parameters of the skew-normal distribution
  params <- cp2dp(c(mean, sd, skew), "SN")

  # generate a sequence of body temperatures and calculate the density
  dsn(temps, xi = params[1], omega = params[2], alpha = params[3])

}

# example usage
tb_distrib <- gen_tb_distrib(
  temps = seq(0, 50, by = 0.1), mean = 25, sd = 5, skew = 0.5
)
tibble(temps = seq(0, 50, by = 0.1), density = tb_distrib) |>
  ggplot(aes(x = temps, y = density)) +
  geom_line(linewidth = 2) +
  labs(x = "Body Temperature (°C)", y = "Density") +
  theme_minimal()
```

Third, I define a function to calculate the expected fitness of a TPC under a given body temperature distribution. I do this using the [Vasseur et al. (2014)](https://royalsocietypublishing.org/rspb/article/281/1779/20132612/77056) approach, which is based on integrating the product of the TPC and the body temperature distribution across all possible body temperatures.

```{r expected-fitness}
#| code-summary: "Defining function to calculate expected fitness"

exp_fitness <- function(topt, ctmin, ctmax, mean_tb, sd_tb, skew_tb) {

  if (sd_tb == 0) {
    tpc(mean_tb, topt, ctmin, ctmax)

  } else {
    # function to calculate integrand
    integrand <- function(temps) {
      tpc(temps, topt, ctmin, ctmax) *  gen_tb_distrib(temps, mean_tb, sd_tb, skew_tb) # nolint
    }

    # integrate
    integrate(integrand, lower = -30, upper = 60)$value
  }
}

# example usage
exp_fitness(topt = 25, ctmin = 10, ctmax = 35, mean_tb = 25, sd_tb = 5, skew_tb = 0.5) # nolint

```

Lastly, I define a function to calculate the risk of suffering a thermal injury, that is experiencing a body temperature below CT<sub>min</sub> or above CT<sub>max</sub>. This function is based on calculating the cumulative density of the body temperature distribution at CT<sub>min</sub> and CT<sub>max</sub>.

```{r thermal-injury-risk}
#| code-summary: "Defining function to calculate thermal injury risk"

ptinj <- function(ctmin, topt, ctmax, mean_tb, sd_tb, skew_tb) {

  if (sd_tb == 0) {
    if (mean_tb < ctmin || mean_tb > ctmax) {
      1
    } else {
      0
    }
  } else {

    # estimate the parameters of the skew-normal distribution
    params <- cp2dp(c(mean_tb, sd_tb, skew_tb), "SN")

    # generate a sequence of body temperatures and calculate the density
    dsn(temps, xi = params[1], omega = params[2], alpha = params[3])

    # calculate ptinj
    ptinj_lower <- psn(ctmin, xi = params[1], omega = params[2], alpha = params[3]) # nolint
    ptinj_upper <- 1 - psn(ctmax, xi = params[1], omega = params[2], alpha = params[3]) # nolint

    # return sum of both tails
    ptinj_lower + ptinj_upper
  }
}

# example usage
ptinj(ctmin = 10, topt = 25, ctmax = 35, mean_tb = 25, sd_tb = 5, skew_tb = 0.5) # nolint

```

## Estimating body temperature distribution parameters from TPC shape

Below, I develop a pseudo-MCMC optimization algorithm to estimate the mean, sd, and skewness of the body temperature distribution that maximizes expected fitness while minimizing the risk of thermal injury. This is just a quick and dirty implementation to explore the idea, so it is not meant to be a rigorous optimization algorithm.

```{r mcmc-optimization}
#| code-summary: "MCMC optimization to estimate body temperature distribution parameters" # nolint

# define TPC shape
topt <- 30
ctmin <- 5
ctmax <- 36

# define number of iterations
n_it <- 4000
n_burn <- 2000

# define objects to store mean_tb, sd_tb, and skew_tb values
mean_tbs <- numeric(n_it)
sd_tbs <- numeric(n_it)
skew_tbs <- numeric(n_it)

# set initial values
mean_tb_init <- topt
sd_tb_init <- 5
skew_tb_init <- 0

# initialize MCMC chain
mean_tbs[1] <- mean_tb_init
sd_tbs[1] <- sd_tb_init
skew_tbs[1] <- skew_tb_init

# run MCMC optimization
for (i in 2:n_it) {

  # propose a new value for mean_tb
  mean_tb_prop <- rnorm(1, mean_tb_init, 5)
  mean_tb_prop <- ifelse(mean_tb_prop < ctmin, ctmin + 0.1, ifelse(mean_tb_prop > ctmax, ctmax - 0.1, mean_tb_prop)) # nolint

  # check if value is accepted
  accept <- rbinom(1, 1, 0.234)
  exp_fit_current <- exp_fitness(topt, ctmin, ctmax, mean_tbs[i - 1], sd_tbs[i - 1], skew_tbs[i - 1]) # nolint
  exp_fit_prop <- exp_fitness(topt, ctmin, ctmax, mean_tb_prop, sd_tbs[i - 1], skew_tbs[i - 1]) # nolint
  ptinj_current <- ptinj(ctmin, topt, ctmax, mean_tbs[i - 1], sd_tbs[i - 1], skew_tbs[i - 1]) # nolint
  ptinj_prop <- ptinj(ctmin, topt, ctmax, mean_tb_prop, sd_tbs[i - 1], skew_tbs[i - 1]) # nolint

  # accept or reject proposal
  if (ptinj_prop < ptinj_current && exp_fit_prop > exp_fit_current && accept == 1) { # nolint
    mean_tbs[i] <- mean_tb_prop
    sd_tbs[i] <- sd_tbs[i - 1]
    skew_tbs[i] <- skew_tbs[i - 1]
  } else {
    mean_tbs[i] <- mean_tbs[i - 1] + rnorm(1, 0, 0.1)
    mean_tbs[i] <- ifelse(mean_tbs[i] < ctmin, ctmin + 0.1, ifelse(mean_tbs[i] > ctmax, ctmax - 0.1, mean_tbs[i])) # nolint
    sd_tbs[i] <- sd_tbs[i - 1] * runif(1, 0.5, 1.5)
    sd_tbs[i] <- ifelse(sd_tbs[i] < 1, 1, sd_tbs[i])
    skew_tbs[i] <- skew_tbs[i - 1] + runif(1, -0.01, 0.01)
    ifelse(skew_tbs[i] < -0.99, skew_tbs[i] <- -0.99, skew_tbs[i])
    ifelse(skew_tbs[i] > 0.99, skew_tbs[i] <- 0.99, skew_tbs[i])
  }

  # propose a new value for sd_tb
  sd_tb_prop <- 1 + rexp(1, 1)

  # check if value is accepted
  accept <- rbinom(1, 1, 0.234)
  exp_fit_current <- exp_fitness(topt, ctmin, ctmax, mean_tbs[i], sd_tbs[i - 1], skew_tbs[i - 1]) # nolint
  exp_fit_prop <- exp_fitness(topt, ctmin, ctmax, mean_tbs[i], sd_tb_prop, skew_tbs[i - 1]) # nolint
  ptinj_current <- ptinj(ctmin, topt, ctmax, mean_tbs[i], sd_tbs[i - 1], skew_tbs[i - 1]) # nolint
  ptinj_prop <- ptinj(ctmin, topt, ctmax, mean_tbs[i], sd_tb_prop, skew_tbs[i - 1]) # nolint

  # accept or reject proposal
  if (ptinj_prop < ptinj_current & exp_fit_prop > exp_fit_current & accept == 1) { # nolint
    mean_tbs[i] <- mean_tbs[i]
    sd_tbs[i] <- sd_tb_prop
    skew_tbs[i] <- skew_tbs[i - 1]
  } else {
    mean_tbs[i] <- mean_tbs[i] + rnorm(1, 0, 0.1)
    mean_tbs[i] <- ifelse(mean_tbs[i] < ctmin, ctmin + 0.1, ifelse(mean_tbs[i] > ctmax, ctmax - 0.1, mean_tbs[i])) # nolint
    sd_tbs[i] <- sd_tbs[i - 1] * runif(1, 0.5, 1.5)
    sd_tbs[i] <- ifelse(sd_tbs[i] < 1, 1, sd_tbs[i])
    skew_tbs[i] <- skew_tbs[i - 1] + runif(1, -0.01, 0.01)
    ifelse(skew_tbs[i] < -0.99, skew_tbs[i] <- -0.99, ifelse(skew_tbs[i] > 0.99, skew_tbs[i] <- 0.99, skew_tbs[i])) # nolint
  }

  # propose a new value for skew_tb
  skew_tb_prop <- runif(1, -0.99, 0.99)

  # check if value is accepted
  accept <- rbinom(1, 1, 0.234)
  exp_fit_current <- exp_fitness(topt, ctmin, ctmax, mean_tbs[i], sd_tbs[i], skew_tbs[i - 1]) # nolint
  exp_fit_prop <- exp_fitness(topt, ctmin, ctmax, mean_tbs[i], sd_tbs[i], skew_tb_prop) # nolint
  ptinj_current <- ptinj(ctmin, topt, ctmax, mean_tbs[i], sd_tbs[i], skew_tbs[i - 1]) # nolint
  ptinj_prop <- ptinj(ctmin, topt, ctmax, mean_tbs[i], sd_tbs[i], skew_tb_prop) # nolint

  # accept or reject proposal
  if (ptinj_prop < ptinj_current && exp_fit_prop > exp_fit_current && accept == 1) { # nolint
    mean_tbs[i] <- mean_tbs[i]
    sd_tbs[i] <- sd_tbs[i]
    skew_tbs[i] <- skew_tb_prop
  } else {
    mean_tbs[i] <- mean_tbs[i - 1] + rnorm(1, 0, 0.1)
    mean_tbs[i] <- ifelse(mean_tbs[i] < ctmin, ctmin + 0.1, ifelse(mean_tbs[i] > ctmax, ctmax - 0.1, mean_tbs[i])) # nolint
    sd_tbs[i] <- sd_tbs[i - 1] * runif(1, 0.5, 1.5)
    sd_tbs[i] <- ifelse(sd_tbs[i] < 1, 1, sd_tbs[i])
    skew_tbs[i] <- skew_tbs[i - 1] + runif(1, -0.01, 0.01)
    ifelse(skew_tbs[i] < -0.99, skew_tbs[i] <- -0.99, ifelse(skew_tbs[i] > 0.99, skew_tbs[i] <- 0.99, skew_tbs[i])) # nolint
  }
}
```

Next, I plot the MCMC chains and the posterior distributions of the mean, sd, and skewness parameters of the body temperature distribution.

```{r plot-chains-and-posteriors}
#| code-summary: "Plotting MCMC chains and posterior distributions"
#| fig-width: 6
#| fig-height: 2
#| out-width: "75%"
#| fig-align: "center"

# plot posteriors of mean_tb, sd_tb, and skew_tb
tibble(mean_tb = mean_tbs, sd_tb = sd_tbs, skew_tb = skew_tbs) |>
  mutate(iteration = row_number()) |>
  pivot_longer(cols = -c(iteration), names_to = "parameter", values_to = "value") |> # nolint
  mutate(parameter = case_when(
    parameter == "mean_tb" ~ "Mean Body Temperature",
    parameter == "sd_tb" ~ "SD of Body Temperature",
    parameter == "skew_tb" ~ "Skewness of Body Temperature"
  )) |>
  ggplot(aes(x = iteration, y = value)) +
  geom_line() +
  facet_wrap(~ parameter, scales = "free") +
  labs(x = "Iteration", y = "Value") +
  theme_minimal()

# plot posterior distributions of mean_tb, sd_tb, and skew_tb
tibble(mean_tb = mean_tbs, sd_tb = sd_tbs, skew_tb = skew_tbs) |>
  mutate(iteration = row_number()) |>
  filter(iteration > n_burn) |> # discard burn-in
  select(-iteration) |>
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") |> # nolint
  mutate(parameter = case_when(
    parameter == "mean_tb" ~ "Mean Body Temperature",
    parameter == "sd_tb" ~ "SD of Body Temperature",
    parameter == "skew_tb" ~ "Skewness of Body Temperature"
  )) |>
  ggplot(aes(x = value)) +
  geom_density(fill = "lightblue") +
  facet_wrap(~ parameter, scales = "free") +
  labs(x = "Value", y = "Density") +
  theme_minimal()
```

In many of the test runs I did, the MCM algorithm swings too much, specially when estimating the mean. I have to think about how to make this as efficient as possible. 

And finally, in the classic Bayesian fashion, I summarize the posterior distributions of the mean, sd, and skewness parameters of the body temperature distribution by calculating their means and 95% confidence intervals.

```{r table-posteriors}
#| code-summary: "Summarizing posterior distributions of body temperature distribution parameters" # nolint

# get average and 95% confidence interval of tb distribution parameters
posterior <- tibble(mean_tb = mean_tbs, sd_tb = sd_tbs, skew_tb = skew_tbs) |>
  mutate(iteration = row_number()) |>
  filter(iteration > n_burn) |> # discard burn-in
  select(-iteration) |>
  pivot_longer(
    cols = everything(), names_to = "parameter", values_to = "value"
  ) |>
  group_by(parameter) |>
  summarise(
    mean = mean(value),
    lower = quantile(value, 0.025),
    upper = quantile(value, 0.975)
  )
kbl(
  posterior |>
    rename(
      `Parameter` = parameter,
      `Mean` = mean,
      `95% CI Lower` = lower,
      `95% CI Upper` = upper
    )
) |>
  kable_styling(full_width = FALSE)
```

Just to get a sense of the predicted body temperature distribution, I plot the TPC and the predicted body temperature distribution based on the mean values of the parameters estimated by the MCMC algorithm.

```{r plot-tpc-and-tb-distribs}
#| code-summary: "Plotting TPC and predicted body temperature distribution"
#| fig-width: 4
#| fig-height: 3
#| out-width: "75%"
#| fig-align: "center"

# plot predicted body temperature distribution and TPC
temps <- seq(0, 50, by = 0.1)
tibble(
  temps = temps,
  tpc_values = tpc(temps, topt, ctmin, ctmax),
  tb_density_mean = gen_tb_distrib(
    temps, posterior$mean[1], posterior$mean[2], posterior$mean[3]
  )
) |>
  ggplot(aes(x = temps)) +
  geom_area(aes(y = tb_density_mean * 2), fill = "red", alpha = 0.5) +
  geom_line(aes(y = tpc_values), color = "black", linewidth = 1.5) +
  geom_vline(xintercept = topt, linetype = "dashed", color = "black") +
  geom_vline(
    xintercept = posterior$mean[1], linetype = "dashed", color = "red"
  ) +
  theme_minimal() +
  labs(x = "Body Temperature (°C)", y = "Performance / Density") +
  coord_cartesian(xlim = c(ctmin - 5, ctmax + 5))
```

## Next steps

- I have to improve the MCMC algorithm so it doesn't bounce as much. I should do some deep diving into the MCMC literature to see how to do this in a more efficient way.
- I should also think about how to incorporate bimodal body temperature distributions or body temperature distributions arising from a function of body temperature fluctuation over time (e.g., diurnal vs nocturnal temperatures). 