---
title: "Plotting maps with iNaturalist and temperature data"
description: "Some quick examples of how to download and plot iNaturalist and temperature maps in R"
date: 2024-06-17
categories: [iNaturalist, mapping]
format:
  html:
    code-fold: true
    out-width: "75%"
    fig-align: "center"
    fig-dpi: 300
    lightbox: true
---

```{r setup}
#| summary: "Setup"
#| message: false
#| warning: false

# packages
library(tidyverse)
library(rinat)
library(geodata)
library(elevatr)
library(sf)
library(terra)
library(tidyterra)
```

## Context

I often find myself needing to quickly download and plot iNaturalist data for a given species, and I also often want to plot temperature maps for a given region. Below is a quick example of how to do so for **western fence lizards** in the area in and around Reno, Nevada. 

## Downloading iNaturalist data

I will use the `get_inat_obs()` function from the `rinat` package to download iNaturalist observations of western fence lizards in the Reno area. Note that `get_inat_obs()` has a lot of options for filtering the data and can be pretty slow if you are downloading a lot of data. 

```{r download-inat-obs}
#| summary: "Downloading iNaturalist observations"

# define bound box for Reno area
reno_bbox <- c(39.25, -120.5, 39.75, -119.5)

# download iNaturalist observations of western fence lizards in the Reno area
lizard_obs <- get_inat_obs(
  taxon_name = "Sceloporus occidentalis",
  quality = "research", # iNat research quality only
  bounds = reno_bbox, # must be S,W,N,E order
  maxresults = 2000 # max can only be 10K
)
```

To have a pretty background, I will also download elevation data from the same area using the `get_elev_raster()` function from the `geodata` package.

```{r download-elevation-data}
#| summary: "Downloading elevation data"

# define grid for elevation data
elev_grid <- expand.grid(
  lat = seq(reno_bbox[1], reno_bbox[3], by = 0.01),
  lon = seq(reno_bbox[2], reno_bbox[4], by = 0.01)
)

# set elevations ar sf object
elev_grid_sf <- st_as_sf(elev_grid, coords = c("lon", "lat"), crs = 4326)

# download elevation data as a raster
elev_raster <- elevatr::get_elev_raster(elev_grid_sf, z = 8, clip = "bbox")

# transform elevation raster to data frame
elev_df <- as.data.frame(elev_raster, xy = TRUE) |> rename(elevation = 3)

# remove negative elevations (i.e., water)
elev_df <- elev_df |> filter(elevation >= 0)
```

Now let's plot the data. Note that I am using the `scale_fill_whitebox_c()` function from `terra` to obtain a nice color palette for elevation data. Check more elevation color palettes [here](https://dieghernan.github.io/tidyterra/reference/scale_whitebox.html?q=atlas#ref-examples). 

```{r plot-inat-elevation}
#| summary: "Plotting iNaturalist observations and elevation data"
#| warning: false
#| message: false
#| fig-width: 7
ggplot() +
  geom_raster(data = elev_df, aes(x = x, y = y, fill = elevation)) +
  geom_contour(
    data = elev_df, aes(x = x, y = y, z = elevation),
    color = "white", alpha = 0.5
  ) +
  scale_fill_whitebox_c(palette = "high_relief") +
  geom_point(
    data = lizard_obs, aes(x = longitude, y = latitude), color = "darkred"
  ) +
  scale_x_continuous(
    expand = c(0, 0), 
    breaks = seq(-120.5, -119.5, by = 0.25),
    labels = paste(seq(-120.5, -119.5, by = 0.25), "°W")
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    breaks = seq(39.25, 39.75, by = 0.25),
    labels = paste(seq(39.25, 39.75, by = 0.25), "°N")
  ) +
  labs(fill = "Elevation (m)") +
  coord_map(
    xlim = c(reno_bbox[2], reno_bbox[4]),
    ylim = c(reno_bbox[1], reno_bbox[3])
  ) +
  theme(axis.title = element_blank())
```

## Downloading and plotting temperature data

Next, I will use the `geodata` package to download temperature data for the same area. I will use the `worldclim_country()` function to download  climate data at a resolution of 0.5 minutes for the US, and then I will crop the data to the Reno area using the `terra` package.

>Note: Downloaidng global climate data can take a while and will be a big file. Set the `path = tempdir()` argument to download the data to a temporary directory that will be deleted after the R session ends. If you want to keep the data, set `path` to a directory on your computer where you want to save the data.

```{r download-temperature-data}
#| eval: FALSE
#| summary: "Downloading temperature data"

# Note: Not running this here because it takes a while to download data.

# download global climate data at 1 minute resolution
climate_data <- geodata::worldclim_country(
  var = "tavg", # average temperature
  res = 0.5, # resolution in minutes (0.5 = 30 seconds)
  country = "usa", # country to download data for
  path = tempdir(), # download to temporary directory
)

# crop climate data to Reno area
climate_data_cropped <- terra::crop(
  climate_data,
  terra::ext(reno_bbox[2], reno_bbox[4], reno_bbox[1], reno_bbox[3]) # must be W,E,S,N order # nolint
)

# get only data for June (i.e., month 6)
climate_data_cropped <- climate_data_cropped[[6]]

# transform climate data to data frame
climate_df <- as.data.frame(climate_data_cropped, xy = TRUE) |>
  rename(temperature = 3)

# save climate_df to .rda file
save(climate_df, file = "posts/inat-and-temp-data/climate_df.rda")
```

Now let's plot the temperature data.

```{r plot-temperature-data}
#| summary: "Plotting temperature data"
#| warning: false
#| message: false
#| fig-width: 7

# load saved climate data
load("climate_df.rda")

# plot temperature data
climate_df |>
  ggplot(aes(x = x, y = y, fill = temperature)) +
  geom_raster() +
  scale_fill_whitebox_c(palette = "muted") +
  scale_x_continuous(
    expand = c(0, 0),
    breaks = seq(-120.5, -119.5, by = 0.25),
    labels = paste0(seq(-120.5, -119.5, by = 0.25), "°W")
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    breaks = seq(39.25, 39.75, by = 0.25),
    labels = paste(seq(39.25, 39.75, by = 0.25), "°N")
  ) +
  labs(fill = "Avg. June\nTemperature (°C)") +
  coord_map(
    xlim = c(reno_bbox[2], reno_bbox[4]),
    ylim = c(reno_bbox[1], reno_bbox[3])
  ) +
  theme(axis.title = element_blank())
```
